\documentclass[11pt]{article}

% Draft ACL long-paper manuscript. Replace this preamble with the official
% ACL style files before submission, e.g., \usepackage[review]{acl}.
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{url}
\usepackage{hyperref}

\newcommand{\yesprob}{p_{\mathrm{yes}}}

\title{Evaluating the Explanatory Validity of Structured LLM Rationales for Forecasting}
\author{Anonymous ACL submission}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Accurate forecasting requires models to provide not only well-calibrated probabilities, but also explanations that make their reasoning inspectable under temporal uncertainty.
We study whether structured rationale prompts improve the forecasting behavior and explanatory quality of large language models (LLMs) on binary future-event questions.
We build a reproducible evaluation pipeline over 1,580 resolved Metaculus-style forecasting questions paired with resolution criteria and news evidence, and compare a neutral baseline against eight rationale structures: predicted event, key attribute, reasoning type, credibility, key conditions, step-by-step reasoning, uncertainty language, and temporal anchors.
Across Qwen2.5-7B-Instruct, Qwen3-32B, and GPT-OSS-120B, we evaluate accuracy, Brier score, expected calibration error, and rationale quality under two automatic LLM judges.
Our results show that structured rationales do not uniformly improve forecasting: temporal anchors and credibility framing can improve some models, while uncertainty-language and key-condition prompts often degrade calibration or accuracy.
Automatic rationale scores correlate with forecast correctness, with plausibility and completeness emerging as the strongest predictors in a SHAP analysis.
These findings suggest that rationale structure is a consequential design choice, but its benefits depend on the target model and evaluation objective.
\end{abstract}

\section{Introduction}

Forecasting future events informs decisions in policy, finance, public health, and strategic planning.
Large language models (LLMs) are increasingly used to generate forecasts accompanied by natural-language rationales, making them attractive for settings where users need both a probability and an explanation.
Yet this combination raises a central question: \textit{does changing the structure of an LLM's rationale improve the reliability of its forecast, or merely change the surface form of its explanation?}

This question matters because probabilistic forecasting is not evaluated only by whether a model chooses the correct label.
A useful forecaster should also be calibrated: events assigned high confidence should occur more often than events assigned low confidence.
At the same time, a rationale should make the forecast inspectable by identifying the relevant event, evidence, assumptions, uncertainty, and temporal constraints.
These goals need not coincide.
A detailed rationale can support an overconfident or incorrect prediction, while a comparatively terse rationale can accompany a well-calibrated forecast.
For forecasting systems, explanation quality and probabilistic reliability therefore have to be measured jointly rather than assumed to improve together.

\paragraph{Challenges in LLM Forecasting.}
Future-event forecasting is difficult for LLMs because it requires reasoning over incomplete evidence, changing temporal context, and verifiable outcomes that may occur after the evidence was written.
Models can overfit to salient evidence, miss resolution-relevant dates, conflate stated plans with realized outcomes, or express verbal uncertainty that is not aligned with numeric confidence.
Moreover, generated rationales may be post-hoc justifications: they can sound coherent without faithfully reflecting the computation that produced the forecast.
This makes forecasting a useful testbed for studying rationale validity, because every prediction can be externally scored after resolution.

\paragraph{Motivation.}
Our approach is motivated by two complementary literatures.
Research on judgmental and crowd forecasting shows that decomposing a question into assumptions, evidence, time horizons, and uncertainty can improve human forecasts \cite{tetlock2015superforecasting}.
More directly, prior work on forecast summarization proposes representing future-related statements through atomic components such as the predicted event, event attributes, source credibility, conditions, rationale, modality, and timestamp \cite{saha2025wisdom}.
Interpretability research, however, cautions that natural-language explanations should not be equated with faithful reasoning traces \cite{jacovi2020towards}.
We use these components as inspiration for our structured rationale variants and treat rationale structure as an experimental intervention: if a component such as temporal anchoring or key-condition enumeration helps LLM forecasting, this should be visible in accuracy, Brier score, calibration, or independently judged rationale quality.

\paragraph{Our Work.}
We evaluate structured LLM rationales on 1,580 resolved Metaculus-style binary forecasting questions paired with descriptions, resolution criteria, temporal metadata, and retrieved news evidence.
We compare a neutral baseline prompt with eight structured variants targeting predicted event restatement, key attribute specification, reasoning type declaration, source credibility, key conditions, step-by-step reasoning, uncertainty language, and temporal anchors.
For the complete checked-in sweep, we evaluate three target models---Qwen2.5-7B-Instruct, Qwen3-32B, and GPT-OSS-120B---across six temperature settings.
We score forecasts using accuracy, Brier score, and expected calibration error (ECE), and we evaluate rationale quality using two automatic LLM judges that rate plausibility, completeness, source consistency, non-hallucination, informativeness, and conciseness.
Finally, we use a SHAP-based diagnostic analysis to test which judged rationale attributes are most predictive of forecast correctness.

Our results show that rationale structure is consequential but not uniformly beneficial.
Temporal anchors yield the best mean Brier score for GPT-OSS-120B and slightly improve over its neutral baseline.
Credibility framing gives Qwen3-32B its best accuracy and ECE, although not its best Brier score.
For Qwen2.5-7B-Instruct, the neutral baseline remains strongest under accuracy and Brier score, suggesting that smaller or less robust models may not benefit from additional rationale constraints.
Across automatic rationale evaluations, plausibility and completeness are the strongest predictors of forecast correctness, while conciseness contributes little because most generated rationales already satisfy the requested length constraints.

\subsection{Research Questions}

We organize the study around three research questions:

\begin{itemize}
    \item \textbf{RQ1:} How do structured rationale components affect probabilistic forecasting performance, as measured by accuracy, Brier score, and ECE?
    \item \textbf{RQ2:} How do structured rationale components affect judged explanatory quality, including plausibility, completeness, source consistency, non-hallucination, informativeness, and conciseness?
    \item \textbf{RQ3:} Which judged rationale attributes are most associated with forecast correctness?
\end{itemize}

\subsection{Hypotheses}

Inspired by the future-related statement components discussed in forecast-summarization research \cite{saha2025wisdom}, we test eight component-level hypotheses corresponding to the eight structured rationale variants beyond the neutral baseline:

\begin{itemize}
    \item \textbf{H1: Predicted event.} Explicitly restating the predicted event improves alignment between the forecast, the resolution criterion, and the generated rationale.
    \item \textbf{H2: Key attribute.} Identifying the key attribute improves specificity by focusing the rationale on the decisive time, quantity, or actor.
    \item \textbf{H3: Reasoning type.} Declaring the reasoning type exposes systematic differences among speculative, plan-based, expert-forecast-based, and trend-based forecasts.
    \item \textbf{H4: Credibility.} Focusing on source credibility improves evidence grounding and reduces unsupported or weakly supported claims.
    \item \textbf{H5: Key conditions.} Enumerating key conditions improves causal precision by making the assumptions behind the forecast explicit.
    \item \textbf{H6: Step-by-step reasoning.} Step-by-step reasoning improves transparency, but may also introduce redundant or formulaic rationales.
    \item \textbf{H7: Uncertainty language.} Explicit uncertainty language improves perceived interpretability and may improve alignment between verbal uncertainty and numeric confidence.
    \item \textbf{H8: Temporal anchors.} Temporal anchors improve temporal grounding by making resolution-relevant dates and time ranges explicit.
\end{itemize}

\subsection{Contributions}

This work makes four contributions:

\begin{itemize}
    \item We introduce a reproducible framework for evaluating structured LLM rationales on resolved binary forecasting questions with evidence, resolution criteria, and temporal metadata.
    \item We provide a controlled comparison of a neutral forecasting prompt and eight structured rationale prompts across three fully evaluated LLMs and six temperature settings.
    \item We jointly evaluate forecast reliability and rationale quality using accuracy, Brier score, ECE, two automatic LLM judges, and feature-attribution analysis.
    \item We show that structured rationales produce model- and metric-dependent trade-offs: temporal and credibility framing help some settings, while richer explanation prompts do not reliably improve calibration.
\end{itemize}

\paragraph{Calibration and Rationale Faithfulness.}
Throughout this paper, calibration is used as an operational measure of probabilistic reliability, not as direct evidence that a generated rationale faithfully represents a model's internal reasoning.
Likewise, high judge-rated rationale quality does not imply causal responsibility for the forecast.
Our experiments therefore evaluate the external relationship between rationale structure, forecast outcomes, and explanation quality, while leaving mechanistic faithfulness as an open question.

\section{Related Work}

\paragraph{LLMs for Future-Event Forecasting.}
Recent work has begun to evaluate whether language models can forecast real-world events rather than merely answer static knowledge questions.
Autocast introduced a large benchmark of forecasting questions paired with date-indexed news evidence, showing that forecasting remains difficult for neural models but improves with model scale and relevant retrieved information \cite{zou2022forecasting}.
Retrieval-augmented forecasting systems have since narrowed the gap with competitive human forecasters by searching for evidence, generating forecasts, and aggregating predictions \cite{halawi2024approaching}.
Work on binary forecasting with LLMs further shows that context matters: adding relevant news summaries can improve performance, while additional prompt scaffolding such as few-shot examples may not always help \cite{mutschlechner2025context}.
These studies establish LLM forecasting as a viable but fragile setting.
Our work differs by holding the task format fixed and asking how the structure of the generated rationale itself affects forecast accuracy, calibration, and explanation quality.

\paragraph{Forecast Structure and Crowd Wisdom.}
Human and crowd forecasting research emphasizes decomposition: strong forecasts often separate the target event, evidence, assumptions, base rates, time horizons, and uncertainty \cite{tetlock2015superforecasting}.
Forecast-summarization work makes a related point from an information-extraction perspective, proposing that future-related statements can be represented through components such as the predicted event, event attributes, source credibility, conditions, rationale, modality, and timestamp \cite{saha2025wisdom}.
This component view directly motivates our prompt variants.
Rather than aggregating many human forecasts, we test whether prompting a single LLM to expose these components changes its probabilistic forecasts or the quality of its rationales.

\paragraph{Calibration and Uncertainty.}
Forecasts are only useful when their probabilities can be trusted.
Proper scoring rules such as the Brier score evaluate the quality of probabilistic predictions \cite{brier1950verification}, while calibration metrics such as expected calibration error assess the alignment between confidence and empirical correctness \cite{guo2017calibration}.
For LLM forecasting, this distinction is especially important because verbal uncertainty and numeric confidence can diverge.
A rationale may say that an event is ``likely'' while assigning an overconfident probability, or it may hedge linguistically without improving calibration.
We therefore evaluate uncertainty language as an explicit rationale intervention rather than assuming that more hedging produces more reliable probabilities.

\paragraph{Rationales and Structured Reasoning.}
Structured prompting methods such as chain-of-thought and self-consistency can improve performance on some reasoning tasks by encouraging intermediate reasoning steps \cite{wei2022chain,wang2023selfconsistency}.
However, explanation research cautions that natural-language rationales may be plausible without being faithful to the model's internal decision process \cite{jacovi2020towards,wiegreffe2019attention}.
This distinction is central for forecasting: a rationale can be fluent, detailed, and source-aware while still supporting an incorrect or miscalibrated forecast.
Accordingly, we evaluate rationales along fine-grained attributes such as plausibility, completeness, source consistency, non-hallucination, informativeness, and conciseness, and then test how those attributes relate to resolved forecast correctness.
The contribution is not another general reasoning prompt, but a controlled evaluation of which forecast-specific rationale components are associated with reliable probabilistic prediction.

\section{Task and Data}

We evaluate binary forecasting questions with resolved yes/no answers.
Each example contains a question, description, resolution criteria, temporal metadata, categories, and up to three retrieved news articles with summaries, keywords, Forecast Relevance Score (FRS), and credibility scores.
The current dataset contains 1,580 resolved questions: 556 resolve to \textsc{yes} and 1,024 resolve to \textsc{no}.
The average number of associated news articles is 2.14, with a maximum of 3.

For each question, a model must output a JSON object containing a predicted answer, a confidence score in $[0,1]$, and a short rationale.
For structured variants, the JSON schema also requires one additional field, such as \texttt{temporal\_anchors} or \texttt{key\_conditions}.
All prompts instruct models to use only the provided question, description, resolution criteria, and evidence summaries.

\section{Prompt Variants}

Table~\ref{tab:variants} summarizes the nine prompt conditions.
The neutral baseline asks for only an answer, confidence, and rationale.
The other eight variants add one targeted reasoning component.

\begin{table}[t]
\centering
\small
\begin{tabular}{ll}
\toprule
Variant & Required rationale structure \\
\midrule
V0 & Neutral baseline \\
V1 & Concrete predicted event \\
V2 & Key attribute: time, quantity, or actor \\
V3 & Reasoning type: speculation, plan, expert forecast, trend \\
V4 & Evidence credibility and recency \\
V5 & Two to four key conditions \\
V6 & Two to three numbered reasoning steps \\
V7 & Explicit uncertainty language \\
V8 & Relevant dates or time ranges \\
\bottomrule
\end{tabular}
\caption{Prompt variants used in the forecasting sweep.}
\label{tab:variants}
\end{table}

\section{Models and Inference}

The checked-in complete sweep covers three target models: Qwen2.5-7B-Instruct, Qwen3-32B, and GPT-OSS-120B.
The repository also contains configuration hooks for additional hosted models, including Llama, DeepSeek, Kimi, Gemma, and GPT-family endpoints, but the quantitative tables in this draft report only models with complete metrics artifacts.

Each model is evaluated across six temperature settings and all nine prompt variants.
Runs write JSON results and metadata containing the provider, model identifier, temperature, output fields, and prompt hashes.
The pipeline supports retry, sharding, resuming incomplete rows, schema validation, and result verification.

\section{Evaluation}

\paragraph{Forecast metrics.}
Let $y \in \{0,1\}$ denote the resolved answer and let $\hat{a}$ be the predicted yes/no label with confidence $c$.
We convert model outputs to a probability of yes:
\[
\yesprob =
\begin{cases}
c, & \hat{a}=\textsc{yes}\\
1-c, & \hat{a}=\textsc{no}.
\end{cases}
\]
We report accuracy, Brier score $\frac{1}{N}\sum_i(\yesprob_i-y_i)^2$, and 10-bin expected calibration error (ECE), where ECE compares predicted-label confidence to empirical correctness within equal-width confidence bins.

\paragraph{Automatic rationale evaluation.}
Two LLM judges, Gemma-4-31B-it and Kimi-K2.5, score rationales using only the supplied question, description, resolution criteria, and evidence digest.
Judges do not score forecast correctness.
They assign six attributes on a 0--1 scale: plausibility, completeness, source consistency, non-hallucination, informativeness, and conciseness.
We average the six attributes to obtain an overall judge score.

\paragraph{Attribution analysis.}
To examine whether judged rationale attributes predict forecast correctness, we fit a classifier from rationale attributes to correctness and compute SHAP feature importances.
This analysis is diagnostic: it does not establish causal effects of rationale quality on correctness.

\section{Results}

\subsection{Temperature Effects}

Table~\ref{tab:temperature} reports the best temperature per model by mean Brier score, averaging over all prompt variants.
The strongest model in the current sweep is GPT-OSS-120B, while both Qwen models perform best at deterministic decoding.

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrrr}
\toprule
Model & Temp. & Acc. & Brier & ECE \\
\midrule
GPT-OSS-120B & 0.25 & 0.817 & 0.166 & 0.121 \\
Qwen2.5-7B-Instruct & 0.00 & 0.734 & 0.200 & 0.112 \\
Qwen3-32B & 0.00 & 0.712 & 0.214 & 0.112 \\
\bottomrule
\end{tabular}
\caption{Best temperature per model by mean Brier score across variants.}
\label{tab:temperature}
\end{table}

\subsection{Prompt Variant Effects}

Table~\ref{tab:variant-best} compares the neutral baseline to the best observed structured variant for each model under different metrics.
The effect of structure is model-dependent.
GPT-OSS-120B benefits most from temporal anchors, which improve mean Brier score from 0.164 to 0.162.
Qwen3-32B obtains its best accuracy and ECE under credibility framing, but this does not improve Brier score.
Qwen2.5-7B-Instruct is strongest under the neutral baseline for accuracy and Brier score, although reasoning-type prompts slightly improve ECE.

\begin{table*}[t]
\centering
\small
\begin{tabular}{llrrr}
\toprule
Model & Condition & Acc. & Brier & ECE \\
\midrule
\multirow{4}{*}{GPT-OSS-120B}
& Baseline & 0.815 & 0.164 & 0.104 \\
& Best Brier: V8 temporal anchors & 0.817 & 0.162 & 0.103 \\
& Best ECE: V5 key conditions & 0.809 & 0.164 & 0.098 \\
& Best accuracy: V8 temporal anchors & 0.817 & 0.162 & 0.103 \\
\midrule
\multirow{4}{*}{Qwen2.5-7B}
& Baseline & 0.743 & 0.196 & 0.104 \\
& Best Brier: V0 baseline & 0.743 & 0.196 & 0.104 \\
& Best ECE: V3 reasoning type & 0.725 & 0.201 & 0.098 \\
& Best accuracy: V0 baseline & 0.743 & 0.196 & 0.104 \\
\midrule
\multirow{4}{*}{Qwen3-32B}
& Baseline & 0.715 & 0.208 & 0.106 \\
& Best Brier: V0 baseline & 0.715 & 0.208 & 0.106 \\
& Best ECE: V4 credibility & 0.730 & 0.211 & 0.086 \\
& Best accuracy: V4 credibility & 0.730 & 0.211 & 0.086 \\
\bottomrule
\end{tabular}
\caption{Prompt-variant effects averaged over temperature runs. Lower Brier and ECE are better.}
\label{tab:variant-best}
\end{table*}

Aggregating across the three models, the neutral baseline has the best mean Brier score (0.189), followed by temporal anchors (0.193) and credibility framing (0.195).
This aggregate result should not be over-interpreted because model-specific effects are heterogeneous.
It does, however, show that adding an explicit rationale component can impose costs as well as benefits.

\subsection{Rationale Quality}

Automatic judge scores do not simply reward longer or more structured outputs.
In the currently available judged runs, the neutral baseline often receives the highest overall judge score, especially for GPT-OSS-120B and both Qwen models.
Step-by-step reasoning and uncertainty-language prompts are usually competitive but do not dominate.
This suggests that concise free-form rationales can be judged as more plausible or source-consistent than forced schemas when the schema encourages formulaic text.

\subsection{Predicting Correctness from Rationale Attributes}

The SHAP analysis in Table~\ref{tab:shap} links automatic rationale attributes to forecast correctness.
Using combined mean judge scores, the attribute classifier obtains ROC-AUC 0.735 and accuracy 0.654 over 13,149 rows.
Plausibility and completeness are the strongest features by mean absolute SHAP value.
Source consistency, informativeness, and non-hallucination differ between correct and incorrect forecasts but have lower attribution magnitude in the fitted model.

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Feature & Mean & Correct--Incorrect & Mean $|\mathrm{SHAP}|$ \\
\midrule
Plausibility & 0.739 & 0.152 & 0.072 \\
Completeness & 0.675 & 0.147 & 0.071 \\
Non-hallucination & 0.658 & 0.126 & 0.021 \\
Informativeness & 0.675 & 0.124 & 0.020 \\
Source consistency & 0.672 & 0.147 & 0.016 \\
Conciseness & 0.908 & 0.024 & 0.012 \\
\bottomrule
\end{tabular}
\caption{Rationale-attribute feature importance for predicting forecast correctness using combined judge scores.}
\label{tab:shap}
\end{table}

\section{Discussion}

The experiments point to three practical lessons.
First, rationale structure should be treated as an intervention, not as a purely cosmetic formatting choice.
The same structural component can improve one metric while degrading another.
Second, temporal and evidential grounding are especially important for forecasting, but they do not guarantee better calibration for every model.
Third, explanation quality and forecast quality are separable: automatic judges can identify rationale attributes associated with correctness, yet high-scoring rationales can still support wrong answers.

The results also caution against relying on explicit uncertainty language as a proxy for calibration.
Prompting models to include words such as ``likely'' or ``possible'' may improve the surface interpretability of rationales, but the checked-in metrics do not show reliable calibration gains from this variant.
In some cases, uncertainty language increases ECE or Brier score, suggesting a mismatch between verbal hedging and numeric confidence.

\section{Limitations}

This draft reports the complete metrics currently available in the repository: three target models, nine prompt variants, and six temperature settings.
The model configuration includes additional systems, but incomplete or partial runs should not be presented as final results until verified.
The rationale-quality evaluation is automatic rather than human-annotated, and LLM judges may share biases with the models being evaluated.
The dataset is drawn from Metaculus-style resolved questions with retrieved news evidence, so results may not generalize to domains where evidence is private, highly technical, or adversarial.
Finally, SHAP analysis is correlational and should not be interpreted as proving that making a rationale more plausible or complete will causally improve forecast accuracy.

\section{Ethical Considerations}

Forecasting systems can influence decisions in finance, policy, public health, and security.
Poorly calibrated forecasts or persuasive but unsupported rationales may mislead users.
Our framework is intended for evaluation and diagnostic use, not as evidence that LLM forecasts should be deployed without human oversight.
The dataset contains public forecasting questions and article-derived evidence, but future releases should document source licenses and remove any sensitive content that appears in retrieved articles.

\section{Conclusion}

We introduced a reproducible evaluation framework for structured LLM rationales in binary forecasting.
Across 1,580 resolved questions, three target models, and nine prompt conditions, we find that rationale structure has heterogeneous effects on accuracy, calibration, and judged explanation quality.
Temporal anchors and credibility framing can help specific models and metrics, but neutral rationales remain strong baselines.
Automatic rationale attributes, especially plausibility and completeness, are predictive of forecast correctness, supporting their use as diagnostic signals.
Future work should add human annotation, expand the model set, and test whether adaptive prompting can select the right rationale structure for each question.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
